Imperative Programming with Variational Effects

### 2018-08-11 paper/paper.tex
edit up to chapter 3

During development of the Resource DSL, we encountered a number of
difficulties in the interaction of variation and side effects. Previous variational
languages have encountered similar issues and have developed a range of
techniques to address them. This thesis builds upon this work
and provides a formal operational semantics for a simple imperative language
that includes both variation and side effects. After presenting this formalization,
we demonstrate how it is successfully applied in our work on the Resource DSL.
}
\acknowledgements{
First and foremost I want to thank my wife, Sara, without whom this certainly
would not have been possible. Thank you for supporting me while I finished
school, and especially for taking such good care of Henry and me. It's been
a long road to get here and I'm glad you were there with me. I love you and
Henry and I'm happy that with the conclusion of my studies we'll have more
time to spend together.

@@

TODO: grant
}

@@

Variation is an extremely common element in modern software. Variation in software can be
thought of as broadly belonging to several interrelated categories: variation that is resolved statically, variation that is preserved during analysis,
and variation that is present during execution.
Variation that is resolved staticially is commonly found in preprocessor directives, command line options, or aspect- or feature-oriented
programming languages \cite{apel2008feature,kiczales1997aspect,prehofer1997feature}.
These all share the common behavior of taking some set of features as input and
producing a program or executable uniquely determined by those features as output. Variational analysis seeks to preserve this variation during program analysis.
For example, the Typechef tool preserves the variation present in C preprocessor annotations in order to typecheck massively variational codebases such as
the Linux kernel \cite{kenner2010typechef}.
Variational that is preserved at runtime, or variational execution, occurs whenever a single
program is run multiple times on related inputs, but may also be viewed as running a program with static variation without eliminating that variation first.
Serial execution of the same program over different inputs can be viewed in the abstract
as equivalent to a single execution across a value representing all inputs and producing a similar value representing all outputs.

@@

Of course, the situation is not quite as hopeless as it might first appear. In all these types of variation
there are significant elements that are shared in common between variants. In statically resolved variation this might take the form of pieces of code such as functions
or classes that are shared between multiple variants. When analyzing such programs, significant portions of the analysis will be shared across many variants. In the case of variational execution often computations are performed across several variants.
By exploiting these commonalities what previously seemed impossible often becomes quite efficient. Software systems analysis that takes advantage of variation
in this way---commonly termed \emph{variationally-aware} analysis---has been developed across a number of areas, including typechecking, type inference, and
testing \cite{kenner2010typechef,chen2014extending,thum2014classification}. In turn, and often in support of such variationally-aware analysis tools, systems of variational execution have been developed for a number
of languages, including PHP, Java, and Javascript \cite{varex,varexj,facted}.

@@

Of course, the design and implementation of the Resource DSL was not without its particular challenges. One particularly interesting and difficult aspect of developing
the variational interpreter for the Resource DSL was accounting for side effects in a variational setting. The Resource DSL itself is primarily interested in side effects: programs are
made up primarily of either operations that somehow modify a resource environment or assertions against the current state of that environment. For this reason, the
Resource DSL is implemented as an imperative language, with side-effectful statements forming the basis of programs in the language. The interactions of side effects
like mutable state and throwing exceptions with variation proved to be difficult to get right. Previous work on making imperative languages variational has developed a
number of techniques to handle the interactions between side effects and variation. This thesis seeks to formalize these techniques using a sample big-step operational
semantics for a simple imperative programming language. With the formal underpinnings thus established, we then turn to a discussion of how they are utilized in more
full-featured language like our Resource DSL.

\section{Contributions and Outline}

@@

Chapter 2 (\emph{Background}) provides an overview of the choice calculus, which provides a formal
model for reasoning about variation. It describes one way of implementing variational values in the
choice calculus via formula trees. Finally, several desirable properties that are enabled by or
crucial to the choice calculus are discussed.

@@

In order to formally represent choices in our programs we employ the choice
calculus \cite{ericthesis,erwig2011choice}. The fundamental unit of the choice calculus is
the \emph{choice}, which is a set of values called \emph{alternatives}.
There are many data structures that can be used to model
variational values, but for our purposes we utilize a data structure we refer
to as a formula tree \cite{walkingshaw2014projectional,walkingshaw2014variational}.
Formula trees represent the set of alternatives as a tree of choices, with concrete
values at the leaves of the tree. Each node of the tree is tagged with a formula from a boolean
algebra consisting of boolean literals, variables, negation, conjunction, and disjunction. We call individual
boolean variables the \emph{dimensions} of the choices, while the expressions as a whole we call \emph{conditions}. Notationally we represent these choice trees
via angle brackets. For example, a choice in dimension $A$
between the values $1$ and $2$ is written $\tagtree{A}{1}{2}$.

@@

The \emph{selection} operation on a choice can be used to eliminate conditions. We define a function
\emph{sel} that takes a \emph{selector} and a variational value and eliminates conditions based off
of the selector. For formula trees selectors are also booleans formulas, and can be seen as specifying
a condition that is assumed to hold true. Therefore, for a given condition and
selector, we eliminate the choice and replace it with the left branch if the selector logically implies
the truth of the condition. Similarly, if the selector logically implies the falsehood of the condition, we eliminate the choice
and replace it with the right branch. If neither implication is valid, the choice remains untouched with
both alternatives left intact.

@@

Selection is \emph{synchronized} for a given
condition, meaning that two choices in the same value or expression that share a condition must always
select the same alternative. For example, in the expression $\tagtree{A}{1}{2}+\tagtree{A}{3}{4}$,
the only possible selections are $1+3$ and $2+4$, while $1+4$ and $2+3$ can never occur due
to synchronization.

We say that a choice is \emph{configured} when all conditions
have been eliminated, yielding a plain value without choices. We call these resulting plain values
\emph{variants}. We define a function \emph{conf} which takes a
\emph{configuration} and a variational value and produces a variant. Under the view of dimensions as representing distinct features, a configuration for a formula tree gives
an assignment for each dimension of the variational value being configured.
A simple way to do this is to define a configuration as a list of the features that are present, with all
other features assumed to be absent. In other words, configurations are a list of variables that should
evaluate to true, with all other variables evaluating to false. Configuring a formula tree is then simply evaluating
each condition given the variable assignment. If the condition evaluates to true, then the left variant
is chosen, otherwise the right variant is chosen.

Using boolean formulas as our tags provides several advantages. For example, we can simplify trees like
$\tagtree{B}{1}{\tagtree{A}{1}{2}}$ to the more compact form $\tagtree{(B \vee A)}{1}{2}$. This is just
one of a number of optimizations that are made possible by representing tags as boolean formulas
\cite{walkingshaw2014projectional,hubbard2016formula}. Formula choices are also more convenient
for maintaining a variational context during execution of a program, a fact that will become
more relevant when we present our work on variational imperative programming later in
this work. Despite these advantages, formulas incur some cost by introducing boolean
satisfiability problems into many common operations, which is an NP-complete problem. However, thanks to improvements in the
efficiency of modern SAT solvers, often the worst-case NP-complete performance of these satisfiability
problems can be avoided.

@@

% correctness and the commuting diagram
Another property of choices and variation that we are interested in is \emph{correctness}. In order
to take advantage of the benefits of sharing, we must develop a variability-aware model of execution
that produces variational values as its result. In order to determine whether our variational results
are correct we need a way to relate them back to the individual variants they represent.

Given a function $f : T \rightarrow U$ and its variational counterpart $g : V \rightarrow W$, we say that
the relationship between the two is correct if for some configuration $c$ the following equality holds:
$\mathit{conf}\ c \circ g = f \circ \mathit{conf}\ c$ \cite{hubbard2016formula}. Put simply, configuring the result
of passing a variational value to a variational function should be the same as pre-configuring the input
and the function itself. This can be visualized as the following commuting diagram:

@@

In order to take advantage of the benefits of sharing provided by choices, we must develop variational
models of execution for a given language. We begin by showing that defining variational execution in a pure,
side-effect free evaluation context is a fairly straight-forward exercise before turning to the much more
difficult problem of variational execution with side effects.

>>>>

During development of the Resource DSL, we encountered a number of
challenges in the interaction of variation and side effects. Previous variational
languages have encountered similar issues and have developed a range of
techniques to address them. This thesis builds upon this work
and provides a formal operational semantics for a simple imperative language
that includes both variation and side effects. After presenting this formalization,
we demonstrate how it is successfully applied in our work on the Resource DSL.
}
\acknowledgements{
First and foremost I want to thank my wife, Sara, without whom this certainly
would not have been possible. Thank you for supporting me while I finished
school, and especially for taking such good care of Henry and me. It's been
a long road to get here and I'm glad you were there with me. I love you and
Henry and I'm happy that with the conclusion of my studies we'll have more
time to spend together.

@@

This work was supported by AFRL Contract FA8750-16-C-0044 (via Raytheon BBN Technologies) under the DARPA BRASS program.
}

@@

Variation is an extremely common element in modern software. Variation in software can be
thought of as broadly belonging to several interrelated categories: variation that is resolved statically, variation that is preserved during analysis,
and variation that is present during execution.
Variation that is resolved staticially is commonly found in preprocessor directives, command line options, or aspect- or feature-oriented
programming \cite{apel2008feature,kiczales1997aspect,prehofer1997feature}.
These all share the common behavior of taking some set of features as input and
producing a program or executable uniquely determined by those features as output. Variational analysis seeks to preserve this variation during program analysis.
For example, the Typechef tool preserves the variation present in C preprocessor annotations in order to typecheck massively variational codebases such as
the Linux kernel \cite{kenner2010typechef}.
Variation that is preserved at runtime, or variational execution, occurs whenever a single
program is run multiple times on related inputs, but may also be viewed as running a program with static variation without eliminating that variation first.
Serial execution of the same program over different inputs can be viewed in the abstract
as equivalent to a single execution across a value representing all inputs and producing a similar value representing all outputs.

@@

Of course, the situation is not quite as hopeless as it might first appear. In all these types of variation
there are significant elements that are shared in common between variants. In statically resolved variation this might take the form of pieces of code such as functions
or classes that are shared between multiple variants. When analyzing such programs, significant portions of the analysis will be shared across many variants. In the case of variational execution often the same computations are performed across several variants.
By exploiting these commonalities what previously seemed impossible often becomes quite efficient. Software systems analysis that takes advantage of variation
in this way---commonly termed \emph{variationally-aware} analysis---has been developed across a number of areas, including typechecking, type inference, and
testing \cite{kenner2010typechef,chen2014extending,thum2014classification}. In turn, and often in support of such variationally-aware analysis tools, systems of variational execution have been developed for a number
of languages, including PHP, Java, and Javascript \cite{varex,varexj,facted}.

@@

Of course, the design and implementation of the Resource DSL was not without its particular challenges. One particularly interesting and difficult aspect of developing
the variational interpreter for the Resource DSL was accounting for side effects in a variational setting. The Resource DSL itself is primarily interested in side effects: programs are
made up primarily of either operations that somehow modify a resource environment or assertions against the current state of that environment. For this reason, the
Resource DSL is implemented as an imperative language, with side-effectful statements forming the basis of programs in the language. The interaction of side effects
like mutable state and throwing exceptions with variation proved to be difficult to get right. Previous work on making imperative languages variational has developed a
number of techniques to handle the interactions between side effects and variation. This thesis seeks to formalize these techniques using a sample big-step operational
semantics for a simple imperative programming language. With the formal underpinnings thus established, we then turn to a discussion of how they are utilized in more
full-featured language like our Resource DSL.

\section{Contributions and Outline}
\label{sec:contrib}

@@

Chapter 2 (\emph{Background}) provides an overview of the choice calculus, which provides a formal
model for reasoning about variation. It describes one way of implementing variational values in the
choice calculus via formula trees. Finally, several desirable properties that are enabled by or
key to the choice calculus are discussed.

@@

In order to formally represent choices in our programs we employ the choice
calculus \cite{ericthesis,erwig2011choice}. The fundamental unit of the choice calculus is
the \emph{choice}, which is a set of values called \emph{alternatives}.
There are many data structures that can be used to model
variational values, but for our purposes we utilize a data structure we refer
to as a formula tree \cite{walkingshaw2014projectional,walkingshaw2014variational}.
Formula trees represent the set of alternatives as a tree of choices, with concrete
values at the leaves of the tree. Each node of the tree is tagged with a formula from a boolean
algebra consisting of boolean literals, variables, negation, conjunction, and disjunction. We call individual
boolean variables the \emph{dimensions} of the choices, while the boolean expressions used as tags we call \emph{conditions}. Notationally we represent these choice trees
via angle brackets. For example, a choice in dimension $A$
between the values $1$ and $2$ is written $\tagtree{A}{1}{2}$.

@@

The \emph{selection} operation on a choice can be used to eliminate conditions. We define a function
\emph{sel} that takes a \emph{selector} and a variational value and eliminates conditions based off
of the selector. For formula trees selectors are also booleans formulas, and can be seen as specifying
a condition that is assumed to hold true. Therefore, for a
selector and a choice with a given condition, we eliminate the choice and replace it with the recursively selected left alternative if the selector logically implies
the truth of the condition. Similarly, if the selector logically implies the falsehood of the condition, we eliminate the choice
and replace it with the recursively selected right alternative. If neither implication is valid, the choice remains untouched with
both alternatives left intact.

@@

Selection is \emph{synchronized} for a given
condition, meaning that two choices in the same value or expression with logically equivalent conditions must always
select the same alternative. For example, in the expression $\tagtree{A}{1}{2}+\tagtree{A}{3}{4}$,
the only possible selections are $1+3$ and $2+4$, while $1+4$ and $2+3$ can never occur due
to synchronization.

We say that a variational value is \emph{configured} when all choices
have been eliminated, yielding a plain value without choices. We call these resulting plain values
\emph{variants}. We define a function \emph{conf} which takes a
\emph{configuration} and a variational value and produces a variant. Under the view of dimensions as representing distinct features, a configuration for a formula tree gives
a boolean assignment for each dimension of the variational value being configured.
A simple way to do this is to define a configuration as a list of the features that are present, with all
other features assumed to be absent. In other words, configurations are a list of variables that should
evaluate to true, with all other variables evaluating to false. Configuring a formula tree is then simply evaluating
each condition given the variable assignment. If the condition evaluates to true, then the left variant
is chosen, otherwise the right variant is chosen.

Using boolean formulas as our tags provides several advantages. For example, we can simplify trees like
$\tagtree{B}{1}{\tagtree{A}{1}{2}}$ to the more compact form $\tagtree{(B \vee A)}{1}{2}$. This is just
one of a number of optimizations that are made possible by representing tags as boolean formulas
\cite{walkingshaw2014projectional,hubbard2016formula}. Formula choices are also more convenient
for maintaining a variational context during execution of a program, a fact that will become
more relevant when we present our work on variational imperative programming later in
Chapter \ref{ch:vimp}. Despite these advantages, formulas incur some cost by introducing boolean
satisfiability problems into many common operations, which is an NP-complete problem. However, thanks to improvements in the
efficiency of modern SAT solvers, often the worst-case NP-complete performance of these satisfiability
problems can be avoided.

@@

This sharing can be accomplished by pushing down choices as deeply as possible in variational
values. For example, consider the variational expression $\tagtree{A}{3*2+4}{3*2+8}$. If we push
the choice down as far as possible, we get $3*2+\tagtree{A}{4}{8}$. This identifies the subexpression
$3*2$ as a shared element between variants $A$ and $\neg A$. A variational interpreter of this
language could then safely only compute that subexpression once.

This sharing is enabled but not required by the choice calculus. It is up to the implementor to
ensure that the maximum degree of sharing is achieved. Nevertheless, there are a number of
strategies for systematically simplifying variational values \cite{ericthesis}. The use of formula
trees permits additional simplification operations \cite{hubbard2016formula,walkingshaw2014projectional}.

% correctness and the commuting diagram
Another property of choices and variation that we are interested in is \emph{variation-preservation}. In order
to take advantage of the benefits of sharing, we must develop a variability-aware model of execution
that produces variational values as its result. In order to determine whether our variational results
are correct we need a way to relate them back to the individual variants they represent.
Variation-preservation assures us that each variant is faithfully modeled by the overall variational computation.

Given a function $f : T \rightarrow U$ and its variational counterpart $g : V \rightarrow W$, we say that
the function $g$ is variation-preserving if for some configuration $c$ the following equality holds:
$\mathit{conf}\ c \circ g = f \circ \mathit{conf}\ c$ \cite{hubbard2016formula}. Put simply, configuring the result
of passing a variational value to a variational function should be the same as pre-configuring the input
and the function itself. This can be visualized as the following commuting diagram:

@@

In order to take advantage of the benefits of sharing provided by choices, we must develop variational
models of execution for a given language. We begin by showing that defining variational execution in a pure,
side-effect free evaluation context is a fairly straightforward exercise before turning to the much more
difficult problem of variational execution with side effects.
